Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-11-04 analysis/paper/paper.Rmd
UPDATE REFERENCES JSON
things had gotten out of sync and the built-in RStudio editor citation workflow was causing problems. Now, relying on the {rbbt} package.

---
title: "Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas"
author:
  - Josh M. London:
      email: josh.london@noaa.gov
      institute: [MML]
      correspondence: true
  - Paul B. Conn:
      institute: [MML]
      correspondence: false
  - Stacie K. Hardy:
      institute: [MML]
      correspondence: false
  - Erin L. Richmond:
      institute: [MML]
      correspondence: false
  - Jay M. Ver Hoef:
      institute: [MML]
      correspondence: false
  - Michael F. Cameron:
      institute: [MML]
      correspondence: false
  - Justin Crawford:
      institute: [ADFG]
      correspondence: false
  - Andrew L. Von Duyke:
      institute: [NSB]
      correspondence: false
  - Lori T. Quakenbush:
      institute: [ADFG]
      correspondence: false
  - Peter L. Boveng:
      institute: [MML]
      correspondence: false
institute:
  - MML: Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA Fisheries, Seattle, Washington, USA
  - ADFG: Arctic Marine Mammals Program, Alaska Department of Fish and Game, Fairbanks, Alaska, USA
  - NSB: Department of Wildlife Management, North Slope Borough, Utqiaġvik, Alaska, USA
output: 
  officedown::rdocx_document:
    reference_docx: ../templates/jml_template.docx
    pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
bibliography: references.json
link-citations: true
csl: ../templates/royal-society-open-science.csl
always_allow_html: true
params:
   run_models: FALSE
---

@@

The global climate disruption is causing considerable reduction in Arctic sea ice extent,
volume, and seasonal presence [@kwok2018erl, @meier2014rg, @wang2017dsrpitsio]. These
changes have ripple effects on Arctic organisms, ecosystems, and the human communities who
live in the region [@huntington2020ncc]. Such disruptions are a particular cause of
concern for ice-associated seals (ringed (*Pusa hispida*), bearded (*Erignathus
barbatus*), spotted (*Phoca largha*), and ribbon seals (*Histriophoca fasciata*)) that
depend on spring and early summer sea ice (March-June) in the Bering and Chukchi seas as a
platform for important functions, such as pupping, nursing, breeding, and molting
[@boveng2009a; @boveng2013; @cameron2010a; @kelly2010]. Limited data and large knowledge
gaps complicate predictions about the ultimate effects of changes in sea ice on the
behavior, abundance, and distribution of these seals. Knowledge about evolutionary
constraints on the timing of reproductive and molting behavior is generally lacking, so it
is difficult to predict the readiness with which ice-associated seal species will be able
to adapt to future changes (e.g., by adjusting pupping or molting schedules to earlier
dates or different locales). Additionally, trends in abundance of these species are
largely unknown, so it is difficult to say what effect, if any, declines in sea ice
habitat have had, or will have, on seal densities.

For ice-associated seals, haul out is tightly linked with sea ice. Ribbon seals rest out
of the water almost exclusively on sea ice and are mostly pelagic outside the spring
pupping, breeding, and molting season [@burns1981]. While spotted and bearded seals rest
on beaches during ice-free months, they strongly prefer sea ice as a resting platform
during the spring and early summer [@burns2009]. The remoteness of the Bering and Chukchi
seas mean direct scientific observation of seal behavior is impractical. Bio-logging
devices are, thus, especially useful tools for collecting key information on movement and
haul-out behavior for these species.

Bio-logging records of time spent out of the water provide valuable data for identifying
covariates that explain variation in haul-out behavior. For instance, in the Antarctic,
Bengtson and Cameron [-@bengtson2004pb] relied on bio-logging data to demonstrate greater
haul-out propensity in juvenile crabeater seals (*Lobodon carcinophaga*) than adults, with
highest probabilities in February and at times close to solar noon. In the Arctic, Von
Duyke et al. [-@vonduyke2020ee] used seal-borne, satellite-linked loggers to corroborate
seasonal changes between diurnal and nocturnal haul-out behavior of ringed seals
previously described by Kelly et al. [-@kelly2010] using labor-intensive VHF radio tags
and direct observation. Bengtson et al. [-@bengtson2005pb] documented a higher propensity
for ringed seal basking near solar noon, as did Ver Hoef et al. [-@verhoef2014sm] in an
analysis of bearded, ribbon, and spotted seals using much larger sample sizes. Olnes et
al. [-@olnes2020pb] showed that the proportion of time bearded seals spent hauled out
progressively increased through spring and summer. And, Ver Hoef et al. [-@verhoef2014sm]
found haul-out probabilities increased gradually starting in March and peaking in May and
June for bearded, ribbon, and spotted seals.

While bio-logging of haul-out patterns clearly has utility for description of
ice-associated seals' natural history and ecology, some studies have focused, as we do
here, specifically on quantifying the probability of seals being hauled out as a means of
accounting for seals in the water, and therefore missed, during aerial surveys. These
studies typically use logistic regression-style analyses to estimate the time-specific
probability of being hauled out base on 'wet/dry' data relayed by bio-loggers. In these
models, haul-out probabilities were expressed as a function of predictive covariates, such
as time of day, day-of-year, sex, age class, and environmental conditions (e.g.,
[@reder2003pb], [@bengtson2004pb], [@bengtson2005pb], [@udevitz2009cjz], [@verhoef2014sm],
and [@southwell2008]). However, sample sizes have often been low and insufficient to
permit strong inference about life history and/or seasonal variation in haul-out
probabilities. For instance, Bengtson and Cameron's [-@bengtson2004pb] study included 5
adult and 2 juvenile crabeater seals, while Bengtson et al.'s [-@bengtson2005pb] study was
based on 6 telemetered ringed seals in the Chukchi Sea. These studies were often further
limited by logistical constraints on tagging fieldwork and the attachment duration or
operational life of satellite tags. In this study, we address some of these limitations by
deploying small bio-loggers designed for attachment on a portion of the study individuals.
These devices are designed to collect data through the molt period (when those adhered to
the hair would fall off) and, in some situations, provide multiple years of data.

Ultimately, knowledge of trends in phenology and abundance (or life history surrogates
such as survival and recruitment) will be necessary to make credible quantitative
predictions about the effects of the climate crisis on the abundance and distribution of
Arctic seal populations. Before we can construct a trend, however, we first require a
baseline. Several studies have contributed estimates of the distribution and abundance of
ice-associated seal species in the Arctic using aerial surveys (e.g., [@bengtson2005pb],
[@conn2014mee], and [@verhoef2014sm]). Such studies were conducted over very large areas
and estimation of absolute abundance required making inference about numerous nuisance
processes affecting the observation of seals on ice. This included availability (only
seals basking on ice were available to be counted), detection probability (observers or
automated detection systems may have missed some seals on ice), species misclassification,
and possible disturbance of seals by aircraft [@conn2014mee; @verhoef2014sm].

In this study, we used data from 15 years of bio-logger records in a comprehensive
investigation of haul-out behavior by 3 species of ice-associated seals in the Bering and
Chukchi seas. We focused our effort on bearded, spotted, and ribbon seals and did not
include ringed seals in the analysis. Ringed seals are unique in their use of snow lairs
in the spring months, and they warrant a separate study that includes additional
environmental covariates, such as the onset of snow melt and snow depth. Our goals are
threefold. First, we wish to establish baseline estimates for the chronology of haul-out
behavior in the critical spring season for each species across different age and sex
classes. Second, we wish to refine estimates of haul-out availability corrections for
aerial surveys in order to improve estimates of seal abundance. Previously estimated
availability correction factors (e.g., [@bengtson2005pb], [@conn2014mee], and
[@verhoef2014sm]) accounted for variables such as the time-of-day and day-of-year, but did
not investigate meteorological variables that have been shown to influence haul-out
behavior of walruses [@reder2003pb; @udevitz2009cjz]. Third, we explored the annual
variability in haul-out timing and possible linkage to dramatic changes in the
availability and extent of seasonal sea ice between 2005 and 2020. Our work extends the
scope of previous haul-out analyses, includes the influence of meteorological variability,
and investigates the potential impact of a changing icescape on the behavior of these
species.

@@

In addition to sex and age class, we gathered additional variables that might help explain
variation in haul-out probabilities. These included day-of-year (for seasonal effects) and
local solar hour (for diurnal effects). Solar hour was calculated using the {solaR}
package [@perpinan2012jss] within the R statistical environment [@rcoreteam2019] using the
daily locations. We also compiled several meteorological variables shown to affect
haul-out behavior in other Arctic pinnipeds [@reder2003pb; @udevitz2009cjz; @perry2017].
In particular, we linked locations from satellite tags to key weather values from the
North American Regional Reanalysis (NARR) model produced by the National Centers for
Environmental Prediction [@mesinger2006bams]. The NARR model assimilates observational
data to produce a long-term picture of weather over North America. Numerous weather
variables are made available across the region 8 times daily. For this study, NARR weather
values were subset to the extent of our study area over the Bering and Chukchi Seas at
3-hr intervals based on the native grid resolution of 32 km (1024 sq. km). The following
meteorological variables were interpolated and assigned to seal locations using a bilinear
method: 1) air temperature at 2m above the earth's surface, 2) wind consisting of
northerly and easterly vector components converted to wind speed using the Euclidean norm,
3) barometric pressure at sea level, and 4) precipitation (Table
\@ref(tab:tblCovariates)).

Models for all species include the following effects: day-of-year, solar hour,
temperature, wind speed, barometric pressure, precipitation, wind chill (represented by a
*wind:temperature* interaction [@udevitz2009cjz]), and day-of-year and time of day
interactions to permit diurnal patterns to change throughout the year. Spotted and ribbon
seal models included age-sex class and interactions between day-of-year and age-sex class,
but we omitted these from bearded seal models due to poor representation of age-sex
classes (Table \@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect
(and an interaction with day-of-year), since bearded seals occupy a substantial range and
we were interested in possible differences in the timing of haul-out behavior along a
latitudinal gradient. We omitted the latitudinal effect from ribbon and spotted seal
models since, during the spring, these species are most prevalent near the southern ice
edge in the Bering Sea [@conn2014mee].

@@

To properly account for temporal dependence within a computationally tractable framework,
we used generalized linear mixed pseudo-models (GLMPMs; [@verhoef2010cs]) to model
variation in haul-out behavior as a function of (1) covariate predictors, (2) temporally
autocorrelated random effects, and (3) individual random effects representing
heterogeneity in individual behavior. We used the {glmmLDTS} package [@verhoef2010cs] to
implement GLMPMs in the R programming environment [@rcoreteam2021]. We explored two
different model formulations for our data, and owing to the large number of records, we
fit separate models to bearded, ribbon, and spotted seal data sets. In our first model
formulation, for each species, we fitted a year-independent model that predicted average
haul-out behavior as a function of demographic, environmental, seasonal, and diurnal
effects. Second, for ribbon and spotted seals (which had considerably more data than
bearded seals), we fitted models that included all the effects from the first model, but
also permitted annual variation in haul-out timing. This second class of models was used
to examine whether haul-out patterns varied by year and to determine the annual timing of
apparent peaks in haul-out behavior.

We followed Ver Hoef et al. [-@verhoef2014sm] in using linear, quadratic, and cubic
effects of day-of-year to represent seasonal changes in behavior. However, unlike previous
models for harbor seals [@london2012] and ice-associated seals [@verhoef2014sm], which
treated hour-of-day as a 24-level categorical variable to capture diurnal cycles, we
adopted a continuous formulation based on Fourier series that provides a flexible model
while preserving the inherent circularity needed for time-of-day effects (i.e., hour 0
should be equal to hour 24). It also represents hour-of-day with 6 parameters, which is a
considerable reduction when compared to a 24 parameter variable, especially when fitting
models with interactions between hour-of-day and other variables (e.g., age-sex class,
day-of-year). According to this approach, we used the following specification for
hour-of-day effects:

@@

A typical model fitting exercise would also include a model selection process. However,
AIC (and similar criteria) is not suitable when using pseudo-likelihoods, because
pseudo-data generated in the model fitting process [@verhoef2010cs] differ between models
[@teneyck2017]. After fitting GLMPM models, we instead used "type III" $F$-tests to
calculate $p$-values [@verhoef2010cs] to evaluate model performance and important terms.
We also produced predictions of haul-out behavior as a function of influential predictors
(e.g. solar hour, day-of-year, age-sex). Weather covariates for these predictions were
based on daily or hourly smoothed weather covariate values across the study region. Such
predictions were then used to develop haul-out probability surfaces, explore conditional
effects of weather covariates, and determine annual peaks in haul-out activity. The timing
of peak haul-out behavior was further used to regress against the annual maximum sea ice
extent in the study region.

@@

Unlike previous analyses of seal haul-out data (e.g. [@verhoef2010cs], [@conn2014mee]), we
also investigated the influence of sex-age class on haul-out probabilities for all species
except bearded seals because of low sample size. While both ribbon and spotted seals
exhibited a unimodal diel haul-out pattern generally centered around local solar noon,
there were key differences across species, age, and sex that match expectations given what
we know of their ecological behavior. Spotted seals are known to form triads during the
breeding season [@burns2009, @mcclintock2015] where a female and dependent pup are
accompanied on the ice by a suitor male. The male waits for the female to wean the pup and
enter estrus, and fends off any other males that may be interested. Triad formation results
in both males and females spending a large portion of the day hauled out on ice and a
protracted spring haul-out season for both sexes. Because mating occurs in the water,
females may also be less inclined to venture out for foraging trips while still nursing
the pup. We see this reflected in the predicted haul-out behavior, with both males and
females exhibiting a broad distribution of time out of the water throughout the solar day
and the season. Ribbon seals are not known to form triads and our model predicts a
progression of increased haul-out behavior with females starting earlier in the season
than males. Notably, female ribbon seals spend a large portion of the day in the water
during the pupping period, aligning with the hypothesis that ribbon seal females continue
foraging while nursing. In the case of both ribbon and spotted seals, subadults are the
first to begin consistent haul-out behavior and follow a typical phocid pattern where
subadults molt first as they do not have any reproductive constraints.

We also investigated the influence of environmental covariates on haul-out probabilities,
including wind speed, temperature, barometric pressure, precipitation, and wind chill.
These have been investigated for walrus (e.g. Udevitz et al. [@udevitz2009cjz]) and a few
select studies of ice-associated seals [@hamilton2018, @perry2017]. Ribbon seals seemed to
be the most influenced by weather, with wind, temperature, barometric pressure, and
precipitation all being important components of the model. Spotted seals were most
affected by wind and barometric pressures. And, for bearded seals, the model indicated
wind and temperature had the greatest impact. In general, and as might be expected, seals
were more likely to haul-out when daily temperatures were warmer, winds speeds were lower,
barometric pressure was higher, and precipitation was lower. These weather conditions
are likely most beneficial for energetic considerations and may also provide the best
situation for predator detection. These results highlight the importance for haul-out
behavior models (and subsequent calculations of availability during surveys) to
incorporate weather covariates.

@@

Previous attempts to estimate the abundance of phocid seals from aerial survey data in the
Bering and Chukchi Seas (e.g. [@bengtson2005pb], [@conn2014mee], [@verhoef2014sm]) have
used estimated haul-out probabilities to correct for the proportion of animals that are in
the water and thus unavailable to be counted during aerial surveys. Although several of
these studies allowed haul-out probabilities to vary by day-of-year and time-of-day,
previous correction factors have not accounted for variability between years, different
weather conditions, and the age-sex class of the sample. In this paper, we have shown that
there can be considerable differences in the number of seals hauled out on ice based on
these factors. We recommend that future abundance analyses employ correction factors that
account for these factors. For instance, it is relatively straightforward to obtain
weather reanalysis products for times and locations that are surveyed and to construct a
relevant correction factor based on predictions of GLMPMs. The most challenging element in
developing availability correction factors is with annual variability. It can be difficult
to get a sufficient sample size to estimate year-specific correction factors, particularly
because research teams would likely need to tag seals and conduct aerial surveys
concurrently which requires considerably more personnel and money. One possible suggestion
is to estimate a "shift" parameter within models for aerial survey counts that allow the
peak of haul-out distributions to be adjusted earlier or later in the year based on the
frequency of counts observed over time. Regardless, researchers should anticipate there
being some unmodeled heterogeneity in availability probability present in abundance
estimates obtained from aerial surveys. This may make trend detection difficult, as one
will not know if moderate differences in abundance estimates are attributable to changes
in abundance or changes in haul-out frequencies.

Predictions of absolute haul-out probability in this paper were somewhat different than
those previously reported for these species, especially for bearded seals. For instance,
Ver Hoef et al. [-@verhoef2014sm)] and Conn et al. [-@conn2014mee] used haul-out
correction factors with maximums of 0.66 for bearded seals, 0.62 for ribbon seals, and
0.54 for spotted seals, where maximums corresponded to times near solar noon in mid-late
May. Applying models that ignore age, sex, and year effects, these probabilities were
0.38, 0.72, and 0.60, respectively, under the current analysis framework. Our current
estimates reflect increased sample sizes in terms of number of animals, but also
improvements to the way data were prepared prior to analysis.

>>>>

---
title: "Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas"
author:
  - Josh M. London:
      email: josh.london@noaa.gov
      institute: [MML]
      correspondence: true
  - Paul B. Conn:
      institute: [MML]
      correspondence: false
  - Stacie K. Hardy:
      institute: [MML]
      correspondence: false
  - Erin L. Richmond:
      institute: [MML]
      correspondence: false
  - Jay M. Ver Hoef:
      institute: [MML]
      correspondence: false
  - Michael F. Cameron:
      institute: [MML]
      correspondence: false
  - Justin Crawford:
      institute: [ADFG]
      correspondence: false
  - Andrew L. Von Duyke:
      institute: [NSB]
      correspondence: false
  - Lori T. Quakenbush:
      institute: [ADFG]
      correspondence: false
  - Peter L. Boveng:
      institute: [MML]
      correspondence: false
institute:
  - MML: Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA Fisheries, Seattle, Washington, USA
  - ADFG: Arctic Marine Mammals Program, Alaska Department of Fish and Game, Fairbanks, Alaska, USA
  - NSB: Department of Wildlife Management, North Slope Borough, Utqiaġvik, Alaska, USA
output: 
  officedown::rdocx_document:
    reference_docx: ../templates/jml_template.docx
    pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
bibliography: references.json
link-citations: true
csl: ../templates/proceedings-of-the-royal-society-b.csl
always_allow_html: true
---

@@

The global climate disruption is causing considerable reduction in
Arctic sea ice extent, volume, and seasonal presence [@kwok2018,
@meier2014, @wang2017]. These changes have ripple effects on
Arctic organisms, ecosystems, and the human communities who live in the
region [@]. Such disruptions are a particular cause
of concern for ice-associated seals (ringed (*Pusa hispida*), bearded
(*Erignathus barbatus*), spotted (*Phoca largha*), and ribbon seals
(*Histriophoca fasciata*)) that depend on spring and early summer sea
ice (March-June) in the Bering and Chukchi seas as a platform for
important functions, such as pupping, nursing, breeding, and molting
[@boveng2009; @boveng2013a; @cameron2010; @kelly2010]. Limited data
and large knowledge gaps complicate predictions about the ultimate
effects of changes in sea ice on the behavior, abundance, and
distribution of these seals. Knowledge about evolutionary constraints on
the timing of reproductive and molting behavior is generally lacking, so
it is difficult to predict the readiness with which ice-associated seal
species will be able to adapt to future changes (e.g., by adjusting
pupping or molting schedules to earlier dates or different locales).
Additionally, trends in abundance of these species are largely unknown,
so it is difficult to say what effect, if any, declines in sea ice
habitat have had, or will have, on seal densities.

For ice-associated seals, haul out is tightly linked with sea ice.
Ribbon seals rest out of the water almost exclusively on sea ice and are
mostly pelagic outside the spring pupping, breeding, and molting season
[@burns1981g]. While spotted and bearded seals rest on beaches during
ice-free months, they strongly prefer sea ice as a resting platform
during the spring and early summer [@burns2009]. The remoteness of the
Bering and Chukchi seas mean direct scientific observation of seal
behavior is impractical. Bio-logging devices are, thus, especially
useful tools for collecting key information on movement and haul-out
behavior for these species.

Bio-logging records of time spent out of the water provide valuable data
for identifying covariates that explain variation in haul-out behavior.
For instance, in the Antarctic, Bengtson and Cameron [-@bengtson2004]
relied on bio-logging data to demonstrate greater haul-out propensity in
juvenile crabeater seals (*Lobodon carcinophaga*) than adults, with
highest probabilities in February and at times close to solar noon. In
the Arctic, Von Duyke et al. [-@vonduyke2020] used seal-borne,
satellite-linked loggers to corroborate seasonal changes between diurnal
and nocturnal haul-out behavior of ringed seals previously described by
Kelly et al. [-@kelly2010] using labor-intensive VHF radio tags and
direct observation. Bengtson et al. [-@bengtson2005a] documented a
higher propensity for ringed seal basking near solar noon, as did Ver
Hoef et al. [-@verhoef2014a] in an analysis of bearded, ribbon, and
spotted seals using much larger sample sizes. Olnes et al.
[-@olnes2020] showed that the proportion of time bearded seals spent
hauled out progressively increased through spring and summer. And, Ver
Hoef et al. [-@verhoef2014a] found haul-out probabilities increased
gradually starting in March and peaking in May and June for bearded,
ribbon, and spotted seals.

While bio-logging of haul-out patterns clearly has utility for
description of ice-associated seals' natural history and ecology, some
studies have focused, as we do here, specifically on quantifying the
probability of seals being hauled out as a means of accounting for seals
in the water, and therefore missed, during aerial surveys. These studies
typically use logistic regression-style analyses to estimate the
time-specific probability of being hauled out base on 'wet/dry' data
relayed by bio-loggers. In these models, haul-out probabilities were
expressed as a function of predictive covariates, such as time of day,
day-of-year, sex, age class, and environmental conditions (e.g.,
[@reder2003], [@bengtson2004], [@bengtson2005a], [@udevitz2009],
[@verhoef2014a], and [@southwell2008]). However, sample sizes have
often been low and insufficient to permit strong inference about life
history and/or seasonal variation in haul-out probabilities. For
instance, Bengtson and Cameron's [-@bengtson2004] study included 5
adult and 2 juvenile crabeater seals, while Bengtson et al.'s
[-@bengtson2005a] study was based on 6 telemetered ringed seals in the
Chukchi Sea. These studies were often further limited by logistical
constraints on tagging fieldwork and the attachment duration or
operational life of satellite tags. In this study, we address some of
these limitations by deploying small bio-loggers designed for attachment
on a portion of the study individuals. These devices are designed to
collect data through the molt period (when those adhered to the hair
would fall off) and, in some situations, provide multiple years of data.

Ultimately, knowledge of trends in phenology and abundance (or life
history surrogates such as survival and recruitment) will be necessary
to make credible quantitative predictions about the effects of the
climate crisis on the abundance and distribution of Arctic seal
populations. Before we can construct a trend, however, we first require
a baseline. Several studies have contributed estimates of the
distribution and abundance of ice-associated seal species in the Arctic
using aerial surveys (e.g., [@bengtson2005a], [@conn2014a], and
[@verhoef2014a]). Such studies were conducted over very large areas
and estimation of absolute abundance required making inference about
numerous nuisance processes affecting the observation of seals on ice.
This included availability (only seals basking on ice were available to
be counted), detection probability (observers or automated detection
systems may have missed some seals on ice), species misclassification,
and possible disturbance of seals by aircraft [@conn2014a;
@verhoef2014a].

In this study, we used data from 15 years of bio-logger records in a
comprehensive investigation of haul-out behavior by 3 species of
ice-associated seals in the Bering and Chukchi seas. We focused our
effort on bearded, spotted, and ribbon seals and did not include ringed
seals in the analysis. Ringed seals are unique in their use of snow
lairs in the spring months, and they warrant a separate study that
includes additional environmental covariates, such as the onset of snow
melt and snow depth. Our goals are threefold. First, we wish to
establish baseline estimates for the chronology of haul-out behavior in
the critical spring season for each species across different age and sex
classes. Second, we wish to refine estimates of haul-out availability
corrections for aerial surveys in order to improve estimates of seal
abundance. Previously estimated availability correction factors (e.g.,
[@bengtson2005a], [@conn2014a], and [@verhoef2014a]) accounted
for variables such as the time-of-day and day-of-year, but did not
investigate meteorological variables that have been shown to influence
haul-out behavior of walruses [@reder2003; @udevitz2009]. Third,
we explored the annual variability in haul-out timing and possible
linkage to dramatic changes in the availability and extent of seasonal
sea ice between 2005 and 2020. Our work extends the scope of previous
haul-out analyses, includes the influence of meteorological variability,
and investigates the potential impact of a changing icescape on the
behavior of these species.

@@

In addition to sex and age class, we gathered additional variables that
might help explain variation in haul-out probabilities. These included
day-of-year (for seasonal effects) and local solar hour (for diurnal
effects). Solar hour was calculated using the {solaR} package
[@perpinan2012] within the R statistical environment
[@rcoreteam2021] using the daily locations. We also compiled several
meteorological variables shown to affect haul-out behavior in other
Arctic pinnipeds [@reder2003; @udevitz2009; @perry2017]. In
particular, we linked locations from satellite tags to key weather
values from the North American Regional Reanalysis (NARR) model produced
by the National Centers for Environmental Prediction
[@mesinger2006]. The NARR model assimilates observational data to
produce a long-term picture of weather over North America. Numerous
weather variables are made available across the region 8 times daily.
For this study, NARR weather values were subset to the extent of our
study area over the Bering and Chukchi Seas at 3-hr intervals based on
the native grid resolution of 32 km (1024 sq. km). The following
meteorological variables were interpolated and assigned to seal
locations using a bilinear method: 1) air temperature at 2m above the
earth's surface, 2) wind consisting of northerly and easterly vector
components converted to wind speed using the Euclidean norm, 3)
barometric pressure at sea level, and 4) precipitation (Table
\@ref(tab:tblCovariates)).

Models for all species include the following effects: day-of-year, solar
hour, temperature, wind speed, barometric pressure, precipitation, wind
chill (represented by a *wind:temperature* interaction
[@udevitz2009]), and day-of-year and time of day interactions to
permit diurnal patterns to change throughout the year. Spotted and
ribbon seal models included age-sex class and interactions between
day-of-year and age-sex class, but we omitted these from bearded seal
models due to poor representation of age-sex classes (Table
\@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect
(and an interaction with day-of-year), since bearded seals occupy a
substantial range and we were interested in possible differences in the
timing of haul-out behavior along a latitudinal gradient. We omitted the
latitudinal effect from ribbon and spotted seal models since, during the
spring, these species are most prevalent near the southern ice edge in
the Bering Sea [@conn2014a].

@@

To properly account for temporal dependence within a computationally
tractable framework, we used generalized linear mixed pseudo-models
(GLMPMs; [@verhoef2010a]) to model variation in haul-out behavior as a
function of (1) covariate predictors, (2) temporally autocorrelated
random effects, and (3) individual random effects representing
heterogeneity in individual behavior. We used the {glmmLDTS} package
[@verhoef2010a] to implement GLMPMs. We explored two different model formulations for our
data, and owing to the large number of records, we fit separate models
to bearded, ribbon, and spotted seal data sets. In our first model
formulation, for each species, we fitted a year-independent model that
predicted average haul-out behavior as a function of demographic,
environmental, seasonal, and diurnal effects. Second, for ribbon and
spotted seals (which had considerably more data than bearded seals), we
fitted models that included all the effects from the first model, but
also permitted annual variation in haul-out timing. This second class of
models was used to examine whether haul-out patterns varied by year and
to determine the annual timing of apparent peaks in haul-out behavior.

We followed Ver Hoef et al. [-@verhoef2014a] in using linear,
quadratic, and cubic effects of day-of-year to represent seasonal
changes in behavior. However, unlike previous models for harbor seals
[@london2012] and ice-associated seals [@verhoef2014a], which treated
hour-of-day as a 24-level categorical variable to capture diurnal
cycles, we adopted a continuous formulation based on Fourier series that
provides a flexible model while preserving the inherent circularity
needed for time-of-day effects (i.e., hour 0 should be equal to hour
24). It also represents hour-of-day with 6 parameters, which is a
considerable reduction when compared to a 24 parameter variable,
especially when fitting models with interactions between hour-of-day and
other variables (e.g., age-sex class, day-of-year). According to this
approach, we used the following specification for hour-of-day effects:

@@

A typical model fitting exercise would also include a model selection
process. However, AIC (and similar criteria) is not suitable when using
pseudo-likelihoods, because pseudo-data generated in the model fitting
process [@verhoef2010a] differ between models [@teneyck2018]. After
fitting GLMPM models, we instead used "type III" $F$-tests to calculate
$p$-values [@verhoef2010a] to evaluate model performance and important
terms. We also produced predictions of haul-out behavior as a function
of influential predictors (e.g. solar hour, day-of-year, age-sex).
Weather covariates for these predictions were based on daily or hourly
smoothed weather covariate values across the study region. Such
predictions were then used to develop haul-out probability surfaces,
explore conditional effects of weather covariates, and determine annual
peaks in haul-out activity. The timing of peak haul-out behavior was
further used to regress against the annual maximum sea ice extent in the
study region.

@@

Unlike previous analyses of seal haul-out data (e.g. [@verhoef2010a],
[@conn2014a]), we also investigated the influence of sex-age class
on haul-out probabilities for all species except bearded seals because
of low sample size. While both ribbon and spotted seals exhibited a
unimodal diel haul-out pattern generally centered around local solar
noon, there were key differences across species, age, and sex that match
expectations given what we know of their ecological behavior. Spotted
seals are known to form triads during the breeding season [@burns2009,
@mcclintock2015] where a female and dependent pup are accompanied on the
ice by a suitor male. The male waits for the female to wean the pup and
enter estrus, and fends off any other males that may be interested.
Triad formation results in both males and females spending a large
portion of the day hauled out on ice and a protracted spring haul-out
season for both sexes. Because mating occurs in the water, females may
also be less inclined to venture out for foraging trips while still
nursing the pup. We see this reflected in the predicted haul-out
behavior, with both males and females exhibiting a broad distribution of
time out of the water throughout the solar day and the season. Ribbon
seals are not known to form triads and our model predicts a progression
of increased haul-out behavior with females starting earlier in the
season than males. Notably, female ribbon seals spend a large portion of
the day in the water during the pupping period, aligning with the
hypothesis that ribbon seal females continue foraging while nursing. In
the case of both ribbon and spotted seals, subadults are the first to
begin consistent haul-out behavior and follow a typical phocid pattern
where subadults molt first as they do not have any reproductive
constraints.

We also investigated the influence of environmental covariates on
haul-out probabilities, including wind speed, temperature, barometric
pressure, precipitation, and wind chill. These have been investigated
for walrus (e.g. Udevitz et al. [-@udevitz2009]) and a few select
studies of ice-associated seals [@hamilton2018, @perry2017]. Ribbon
seals seemed to be the most influenced by weather, with wind,
temperature, barometric pressure, and precipitation all being important
components of the model. Spotted seals were most affected by wind and
barometric pressures. And, for bearded seals, the model indicated wind
and temperature had the greatest impact. In general, and as might be
expected, seals were more likely to haul-out when daily temperatures
were warmer, winds speeds were lower, barometric pressure was higher,
and precipitation was lower. These weather conditions are likely most
beneficial for energetic considerations and may also provide the best
situation for predator detection. These results highlight the importance
for haul-out behavior models (and subsequent calculations of
availability during surveys) to incorporate weather covariates.

@@

Previous attempts to estimate the abundance of phocid seals from aerial
survey data in the Bering and Chukchi Seas (e.g. [@bengtson2005a],
[@conn2014a], [@verhoef2014a]) have used estimated haul-out
probabilities to correct for the proportion of animals that are in the
water and thus unavailable to be counted during aerial surveys. Although
several of these studies allowed haul-out probabilities to vary by
day-of-year and time-of-day, previous correction factors have not
accounted for variability between years, different weather conditions,
and the age-sex class of the sample. In this paper, we have shown that
there can be considerable differences in the number of seals hauled out
on ice based on these factors. We recommend that future abundance
analyses employ correction factors that account for these factors. For
instance, it is relatively straightforward to obtain weather reanalysis
products for times and locations that are surveyed and to construct a
relevant correction factor based on predictions of GLMPMs. The most
challenging element in developing availability correction factors is
with annual variability. It can be difficult to get a sufficient sample
size to estimate year-specific correction factors, particularly because
research teams would likely need to tag seals and conduct aerial surveys
concurrently which requires considerably more personnel and money. One
possible suggestion is to estimate a "shift" parameter within models for
aerial survey counts that allow the peak of haul-out distributions to be
adjusted earlier or later in the year based on the frequency of counts
observed over time. Regardless, researchers should anticipate there
being some unmodeled heterogeneity in availability probability present
in abundance estimates obtained from aerial surveys. This may make trend
detection difficult, as one will not know if moderate differences in
abundance estimates are attributable to changes in abundance or changes
in haul-out frequencies.

Predictions of absolute haul-out probability in this paper were somewhat
different than those previously reported for these species, especially
for bearded seals. For instance, Ver Hoef et al. [-@verhoef2014a)] and
Conn et al. [-@conn2014a] used haul-out correction factors with
maximums of 0.66 for bearded seals, 0.62 for ribbon seals, and 0.54 for
spotted seals, where maximums corresponded to times near solar noon in
mid-late May. Applying models that ignore age, sex, and year effects,
these probabilities were 0.38, 0.72, and 0.60, respectively, under the
current analysis framework. Our current estimates reflect increased
sample sizes in terms of number of animals, but also improvements to the
way data were prepared prior to analysis.
